Finding a closest point in a lattice of Voronoi's first kind

### 2014-07-09 paper.tex
minor

\begin{figure}[tp] 
	\centering      
		\includegraphics{figs/latticefigures-1.mps} 
		\caption{The $2$-dimensional lattice with basis vectors $(3,0.6)$ and $(0.6,3)$.  The lattice points are represented by dots and the relevant vectors are circled.  The Voronoi cell $\vor(\Lambda)$ is the shaded region and the packing radius $\rho$ and corresponding sphere packing (circles) are depicted.
}     
		\label{lattices:fig:vorregion}   
\end{figure}

@@

\begin{figure}[tp] 
	\centering      
		\includegraphics{figs/iterativeseriesexample-1.mps} 
		\caption{Example of the iterative procedure described in~\eqref{eq:relvectsminimslicer} to compute a closest lattice point to $y = (4,3.5)$ (marked with a cross) in the $2$-dimensional lattice generated by basis vectors $(2,0.4)$ and $(0.4,2)$.  The initial lattice point for the iteration is $x_0 = (-4.4,-2.8)$.  The shaded region is the Voronoi cell surrounding the closest lattice point $x_6 = (4.4,2.8)$.}       
		\label{lattices:fig:iterativeexample} 
\end{figure}

@@

An $n$-dimensional lattice $\Lambda$ is said to be of \emph{Voronoi's first kind} if it has what is called an \emph{obtuse superbasis}~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}.  That is, there exists a set of $n+1$ vectors $b_1,\dots,b_{n+1}$ such that $b_1,\dots,b_n$ are a basis for $\Lambda$,
\begin{equation}\label{eq:superbasecond}
b_1 + b_2 \dots + b_{n+1} = 0
\end{equation}
(the \emph{superbasis} condition), and the inner products satisfy
\begin{equation}\label{eq:obtusecond}
q_{ij} = b_i \cdot b_j \leq 0, \qquad \text{for} \qquad i,j = 1,\dots,n+1, i \neq j
\end{equation}
(the \emph{obtuse} condition).  The $q_{ij}$ are called the \emph{Selling parameters}~\cite{Selling1874}.  It is known that all lattices in dimensions less than or equal to $3$ are of Voronoi's first kind~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}.  
%BLERG.  THIS MIGHT BETTER GO IN THE GRAPH SECTION
% There is a one-to-one correspondence between the obtuse superbasis of a lattice of Vornoi's first kind and what is called the \emph{Laplacian} of a weighted graph.  To se this, let
% \[
% B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
% \]
% be the matrix with columns given by $b_1,\dots,b_{n+1}$ and define the $n+1$ by $n+1$ \emph{Gram matrix} $Q = B^\prime B$.  The elements of $Q$ are given by the Selling parameters, that is, $Q_{ij} = q_{ij}$.  Let $G$ be a graph with $n+1$ vertices $v_1,\dots,v_{n+1}$ and edges $e_{ij}$ connecting $v_i$ to $v_j$.  To each edge assign the \emph{weight} $w_{ij} = -q_{ij} \leq 0$.  Let $d_{ii} = \sum_{j \neq i} w_{ij}$ be the sum of weights of edges connected to vertex $v_i$. The Laplacian of the graph $G$ is the $n+1$ by $n+1$ matrix with elements
% \[
% L_{ij} = \begin{cases}
% d_{ii} & i = j \\
% -w_{ij} & i \neq j. 
% \end{cases}
% \]
An interesting property of lattices of Voronoi's first kind is that their relevant vectors have a straightforward description.

@@

BLERG: It will be instructive for us to give a proof of this theorem.  In doing so we will introduce some of the notation that will be used in Section~\ref{}.

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$ and let $S$ and $T$ be subsets of the indices of $p$.  The following equalities hold:
\begin{enumerate}
\item  $\|Bp\|^2 - \|B(p + \onebf_S)\|^2 = \Phi(S, p)$, \label{eq:lem:decSellingsinc}
\item  $\|Bp\|^2 - \|B(p - \onebf_S)\|^2 = \Phi(\bar{S}, p)$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + \onebf_S - \onebf_T)\|^2 = \Phi(S, p) + \Phi(\bar{T},p) + 2\sum_{i\in S}\sum_{j\in T} q_{ij} }$.  \label{eq:lem:decSellingsincdec}
\end{enumerate}
\end{lemma}
\begin{proof}
Part~\ref{eq:lem:decSellingsincdec} follows immediately from parts~\ref{eq:lem:decSellingsinc} and~\ref{eq:lem:decSellingsdec} because
\begin{align*}
\|Bp\|^2 - &\|B(p + \onebf_S - \onebf_T)\|^2 \\
&= \|Bp\|^2-\|B(p + \onebf_S)\|^2 +  \|Bp\|^2-\|B(p - \onebf_T)\|^2 + 2\sum_{i\in S}\sum_{j\in T} q_{ij}.
\end{align*}
We give a proof for part~\ref{eq:lem:decSellingsinc}.  The proof for part~\ref{eq:lem:decSellingsdec} is similar.  
Put $Q = B^\prime B$ where superscript $^\prime$ indicates the vector or matrix transpose.  The $n+1$ by $n+1$ matrix $Q$ has elements given by the Selling parameters, that is, $Q_{ij} = q_{ij}=\dotprod{b_i}{b_j}$.  Denote by $\onebf$ the column vector of length $n+1$ containing all ones.  Now $B\onebf = \sum_{i=1}^{n+1}b_i = \zerobf$ as a result of the superbasis condition~\eqref{eq:superbasecond} and so $Q\onebf = \zerobf$.  Since $\onebf_S = \onebf - \onebf_{\bar{S}}$ it follows that $Q\onebf_S = -Q\onebf_{\bar{S}}$.  With $\circ$ the elementwise vector product, i.e., the Schur or Hadamard product, we have
 \begin{align*}
 \|Bp\|^2 - \|B(p + \onebf_S)\|^2 &= -\onebf^\prime_S Q \onebf_S - 2 p^\prime Q \onebf_S \\
 &= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 p^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_S - 2 (p\circ\onebf_S)^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_{S} + 2 (p\circ\onebf_S)^\prime Q \onebf_{\bar{S}}
 \end{align*}
which is precisely $\Phi(S,p)$.
% Observe that
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
% \]
% and
% \[
% \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
% \]
% since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
% \begin{align*}
% \|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
% &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
% &= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
% \end{align*}
% This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - \onebf_S$.  It follows that
% \[
% \|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
% \]
% where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
% \[
% d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
% \]
% If $i \in S$ and $j \notin S$ then
% \[
% d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
% \]
% while if $i \notin S$ and $j \in S$ then
% \[
% d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
% \]
% We thus have
% \begin{align*}
% \|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
% &= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
% \end{align*}
% where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}

>>>>

\begin{figure}[p] 
	\centering      
		\includegraphics{figs/latticefigures-1.mps} 
		\caption{The $2$-dimensional lattice with basis vectors $(3,0.6)$ and $(0.6,3)$.  The lattice points are represented by dots and the relevant vectors are circled.  The Voronoi cell $\vor(\Lambda)$ is the shaded region and the packing radius $\rho$ and corresponding sphere packing (circles) are depicted.
}     
		\label{lattices:fig:vorregion}   
\end{figure}

\begin{figure}[p] 
	\centering      
		\includegraphics{figs/iterativeseriesexample-1.mps} 
		\caption{Example of the iterative procedure described in~\eqref{eq:relvectsminimslicer} to compute a closest lattice point to $y = (4,3.5)$ (marked with a cross) in the $2$-dimensional lattice generated by basis vectors $(2,0.4)$ and $(0.4,2)$.  The initial lattice point for the iteration is $x_0 = (-4.4,-2.8)$.  The shaded region is the Voronoi cell surrounding the closest lattice point $x_6 = (4.4,2.8)$.}       
		\label{lattices:fig:iterativeexample} 
\end{figure}

@@

@@

An $n$-dimensional lattice $\Lambda$ is said to be of \emph{Voronoi's first kind} if it has what is called an \emph{obtuse superbasis}~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}.  That is, there exists a set of $n+1$ vectors $b_1,\dots,b_{n+1}$ such that $b_1,\dots,b_n$ are a basis for $\Lambda$,
\begin{equation}\label{eq:superbasecond}
b_1 + b_2 \dots + b_{n+1} = 0
\end{equation}
(the \emph{superbasis} condition), and the inner products satisfy
\begin{equation}\label{eq:obtusecond}
q_{ij} = b_i \cdot b_j \leq 0, \qquad \text{for} \qquad i,j = 1,\dots,n+1, i \neq j
\end{equation}
(the \emph{obtuse} condition).  The $q_{ij}$ are called the \emph{Selling parameters}~\cite{Selling1874}.  It is known that all lattices in dimensions less than or equal to $3$ are of Voronoi's first kind~\cite{ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}.  
%%%%%%%%%%%%%%%%%%%%
% There is a one-to-one correspondence between the obtuse superbasis of a lattice of Vornoi's first kind and what is called the \emph{Laplacian} of a weighted graph.  To see this, let
% \[
% B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
% be the matrix with columns given by $b_1,\dots,b_{n+1}$ and define the $n+1$ by $n+1$ symmetric \emph{Gram matrix} $Q = B^\prime B$.  The elements of $Q$ are given by the Selling parameters, that is, $Q_{ij} = q_{ij}$.  Because of the superbasis condition the row and column sums of $Q$ are zero.  In other words, $Q\onebf = \zerobf$ and $\onebf^\prime Q = \zerobf^\prime$ where $\onebf$ and $\zerobf$ denote column vectors of ones and zeros respectively.  Let $G$ be a graph with $n+1$ vertices $v_1,\dots,v_{n+1}$ and edges $e_{ij}$ connecting $v_i$ to $v_j$.  To each edge assign the \emph{weight} $w_{ij} = -q_{ij} \leq 0$.  Let $d_{ii} = \sum_{j \neq i} w_{ij}$ be the sum of weights of edges connected to vertex $v_i$. The Laplacian of the graph $G$ is the $n+1$ by $n+1$ matrix with elements
% \[
% L_{ij} = \begin{cases}
% d_{ii} & i = j \\
% -w_{ij} & i \neq j. 
% \end{cases}
%%%%%%%%%%%%%%%%%%%
% Lattices of Vornoi's first kind have some other interesting properties.  Let
% \[
% B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
% be the matrix with columns given by $b_1,\dots,b_{n+1}$ and define the $n+1$ by $n+1$ symmetric \emph{Gram matrix} $Q = B^\prime B$ where superscript $^\prime$ indicates the vector or matrix transpose.  The elements of $Q$ are given by the Selling parameters, that is, $Q_{ij} = q_{ij}$.  Because of the superbasis condition the row and column sums of $Q$ are zero, i.e., 
% \[
% \sum_{i = 1}^n q_{ij} = \sum_{j=1}^n q_{ij} = 0.
% Equivalently, $Q\onebf = \zerobf$ and $\onebf^\prime Q = \zerobf^\prime$ where $\onebf$ and $\zerobf$ denote column vectors of ones and zeros respectively.  The following formula will be useful.  For $x\in\reals^{n+1}$ and $y \in \reals^{n+1}$, the quadratic form
% \begin{align}
% x^\prime Q y &= \frac{1}{2} ( x^\prime Qy + y^\prime Qx ) \nonumber \\ 
% &= \frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij}(x_iy_j + x_jy_i) \nonumber \\
% &= \frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij}\big( x_iy_i + x_jy_j - (x_i - x_j)(y_i - y_j) \big) \nonumber \\
% &= -\sum_{i=1}^{n+1} \sum_{j=i+1}^{n+1} q_{ij}(x_i - x_j)(y_i - y_j). \label{eq:extendedsellings}
% \end{align}
% The first line in the above equation follows from the symmetry of $Q$.  The last line follows because
% \[
% \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij}x_jy_j = \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij}x_iy_i = \sum_{i=1}^{n+1} x_iy_i \sum_{j=1}^{n+1} q_{ij} = 0
% because the row and column sums of $Q$ are zero.  Putting $y = x$ yields what is called~\emph{Selling's formula}~\cite{Selling1874,ConwaySloane1992_voronoi_lattice_3d_obtuse_superbases}\cite[Proposition~2.3.1]{Valentin2003_coverings_tilings_low_dimension},
% \begin{equation}\label{eq:sellings}
% \|Bx\|^2 = x^\prime Q x = -\sum_{i=1}^{n+1} \sum_{j=i+1}^{n+1} q_{ij}(x_i - x_j)^2.
% \end{equation}
%%%%%%%%%%%%%%%%%%%%%
An interesting property of lattices of Voronoi's first kind is that their relevant vectors have a straightforward description.

@@

% \begin{align*}
% x^\prime Q y &= \sum_{i=1}^n \sum_{j=1}^n q_{ij} x_i y_j \\
% &= \sum_{i=1}^n \sum_{j=1}^n q_{ij} x_i y_j  - 2
% \end{align*}
% \[
% (x-y)^\prime Q (x-y) = \sum_{i=1}^n \sum_{j=1}^n q_{ij} (x_i-y_i)^2
% \]

%BLERG: It will be instructive for us to give a proof of this theorem.  In doing so we will introduce some of the notation that will be used in Section~\ref{}.

@@

\begin{lemma}\label{lem:decSellings}
Let $p \in \reals^{n+1}$ and let $S$ and $T$ be subsets of the indices of $p$.  The following equalities hold:
\begin{enumerate}
\item  $\|Bp\|^2 - \|B(p + \onebf_S)\|^2 = \Phi(S, p)$, \label{eq:lem:decSellingsinc}
\item  $\|Bp\|^2 - \|B(p - \onebf_S)\|^2 = \Phi(\bar{S}, p)$, \label{eq:lem:decSellingsdec}
\item  ${\displaystyle \|Bp\|^2 - \|B(p + \onebf_S - \onebf_T)\|^2 = \Phi(S, p) + \Phi(\bar{T},p) + 2\sum_{i\in S}\sum_{j\in T} q_{ij} }$.  \label{eq:lem:decSellingsincdec}
\end{enumerate}
\end{lemma}
\begin{proof}
Part~\ref{eq:lem:decSellingsincdec} follows immediately from parts~\ref{eq:lem:decSellingsinc} and~\ref{eq:lem:decSellingsdec} because
\begin{align*}
\|Bp\|^2 - &\|B(p + \onebf_S - \onebf_T)\|^2 \\
&= \|Bp\|^2-\|B(p + \onebf_S)\|^2 +  \|Bp\|^2-\|B(p - \onebf_T)\|^2 + 2\sum_{i\in S}\sum_{j\in T} q_{ij}.
\end{align*}
We give a proof for part~\ref{eq:lem:decSellingsinc}.  The proof for part~\ref{eq:lem:decSellingsdec} is similar.  
Put $Q = B^\prime B$ where superscript $^\prime$ indicates the vector or matrix transpose.  The $n+1$ by $n+1$ matrix $Q$ has elements given by the Selling parameters, that is, $Q_{ij} = q_{ij}=\dotprod{b_i}{b_j}$.  Denote by $\onebf$ the column vector of length $n+1$ containing all ones.  Now $B\onebf = \sum_{i=1}^{n+1}b_i = \zerobf$ as a result of the superbasis condition~\eqref{eq:superbasecond} and so $Q\onebf = \zerobf$.  Since $\onebf_S = \onebf - \onebf_{\bar{S}}$ it follows that $Q\onebf_S = -Q\onebf_{\bar{S}}$.  With $\circ$ the elementwise vector product, i.e., the Schur or Hadamard product, we have
 \begin{align*}
 \|Bp\|^2 - \|B(p + \onebf_S)\|^2 &= -\onebf^\prime_S Q \onebf_S - 2 p^\prime Q \onebf_S \\
 &= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 p^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_S - 2 (p\circ\onebf_S)^\prime Q \onebf_S \\
&= \onebf^\prime_S Q \onebf_{\bar{S}} - 2 (p\circ\onebf_{\bar{S}})^\prime Q \onebf_{S} + 2 (p\circ\onebf_S)^\prime Q \onebf_{\bar{S}}
 \end{align*}
which is precisely $\Phi(S,p)$.
% % Observe that
% % \[
% % \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_i^2 =  \sum_{i=1}^{n+1}p_i^2\sum_{j=1}^{n+1} q_{ij} = 0
% % \]
% % and
% % \[
% % \sum_{i=1}^{n+1}\sum_{j=1}^{n+1} q_{ij} p_j^2 =  \sum_{j=1}^{n+1}p_j^2\sum_{i=1}^{n+1} q_{ij} = 0
% % \]
% % since $\sum_{i=1}^{n+1} q_{ij} = \sum_{j=1}^{n+1} q_{ij} = 0$ as a result of the superbasis condition~\eqref{eq:superbasecond}.  Now
% % \begin{align*}
% % \|Bp\|^2 &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} p_i p_j \\
% % &= \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (\tfrac{1}{2}p_j^2 + p_i p_j -  \tfrac{1}{2}p_i^2) \\
% % &= -\frac{1}{2} \sum_{i=1}^{n+1} \sum_{j=1}^{n+1} q_{ij} (p_i - p_j)^2.
% % \end{align*}
% % This is closely related to what is known as Sellings formula~\cite[Proposition 2.3.1]{Valentin2003_coverings_tilings_low_dimension}~\cite{Selling1874}.  Put $g = p - \onebf_S$.  It follows that
% % \[
% % \|Bp\|^2 - \|Bg\|^2 = \frac{1}{2} \sum_{i=1}^{n+1}\sum_{j=1}^{n+1}q_{ij}d_{ij},
% % \]
% % where $d_{ij} = (g_i - g_j)^2 - (p_i - p_j)^2$.  If both $i \notin S$ and $j \notin S$ then $p_i = g_i$, $p_j = g_j$ and so $d_{ij} = 0$.  Similarly if both $i \in S$ and $j \in S$ then 
% % \[
% % d_{ij} = (p_i-1 - p_j+1)^2 - (p_i - p_j)^2 = 0.
% % \]
% % If $i \in S$ and $j \notin S$ then
% % \[
% % d_{ij} = (p_i-1 - p_j)^2 - (p_i - p_j)^2= 1 + 2p_j - 2p_i,
% % \]
% % while if $i \notin S$ and $j \in S$ then
% % \[
% % d_{ij} = (p_i - p_j+1)^2 - (p_i - p_j)^2 = 1 + 2p_i - 2p_j.
% % \]
% % We thus have
% % \begin{align*}
% % \|Bp\|^2 - \|Bg\|^2 &= \frac{1}{2} \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i) + \frac{1}{2} \sum_{i \notin S}\sum_{j \in S}q_{ij}(1 + 2p_i - 2p_j) \\
% % &= \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_j - 2p_i),
% % \end{align*}
% % where the last line holds because $q_{ij} = q_{ji}$.
\end{proof}
